---
title: "Report"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    keep_tex: true
header-includes:
  - \usepackage{graphicx}
  - \usepackage{caption}
  - \usepackage{float}
params:
  dat: NULL
  filtering_threshold: NULL
  LOD_method: NULL
  LLOQ_method: NULL
  ULOQ_method: NULL
  LOD_type: NULL
  cv_threshold: NULL
  sample_type_PCA_types: "all"
  sample_type_PCA_threshold: NULL
  sample_type_PCA_variance_threshold: NULL
  sample_type_PCA_variance_max_num: NULL
  sample_type_PCA_variance_cum: TRUE
  group_PCA_threshold: NULL
  group_PCA_variance_threshold: NULL
  group_PCA_variance_max_num: NULL
  group_PCA_variance_cum: TRUE
  filtering_threshold_ex: FALSE
---

<style type="text/css">
.main-container {max-width: 100%;}
.row {display: flex;}
.column {flex: 50%;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(ggplot2)
library(MetaboCrates)
library(dplyr)
library(ggpubr)
```

## Data summary

```{r}
m_num <- length(attr(params[["dat"]], "metabolites"))
s_num <- params[["dat"]] %>%
  filter(`sample type` == "Sample") %>%
  nrow()
sample_types <- pull(attr(params[["dat"]], "samples"),
                     `sample type`)
materials <- unique(pull(params[["dat"]], "material"))
op <- unique(pull(params[["dat"]], "op"))
pb_codes <- unique(pull(params[["dat"]],
                        "plate bar code"))

total_mv_count <-
  attr(params[["dat"]], "NA_info")[["counts"]] %>%
  select(n) %>%
  sum()
```
Dataset contains `r m_num` compounds and `r s_num` samples.

Sample types (`r length(sample_types)`): `r paste0(sample_types, collapse = ", ")`.

Materials (`r length(materials)`): `r paste0(materials, collapse = ", ")`.

OP (`r length(op)`): `r paste0(op, collapse = ", ")`.

Plate bar codes (`r length(pb_codes)`): `r paste0(pb_codes, collapse = ", ")`.

Initial missing values count was `r total_mv_count`.

```{r, fig.height=3, fig.width=6, fig.cap="Counts of missing values by type.", fig.align='center'}
plot_mv_types(params[["dat"]])
```

```{r, results='asis', eval=!is.null(attr(params[["dat"]], "group"))}
lvls_count <- params[["dat"]] %>%
  filter(`sample type` == "Sample") %>%
  group_by(across(all_of(attr(params[["dat"]], "group")))) %>%
  summarise() %>%
  nrow()

cat("## Grouping")
cat("  \n  \nThe data was grouped by *",
    paste0(attr(params[["dat"]], "group"), collapse = ", "),
    "*. The number of unique levels is ",
    lvls_count,
    ".", sep = "")
```

```{r, fig.height=4, fig.width = 5, eval=!is.null(attr(params[["dat"]], "group")), fig.cap="The counts of grouping levels.", fig.pos="H", fig.align='center'}
plot_groups(params[["dat"]])
```

## Compounds filtering

```{r, results='asis', eval=params[["filtering_threshold_ex"]]}
cat("Missing values threshold was set to ",
    params[["filtering_threshold"]],
    "%.  \n   \n", sep = "")

LOD_to_remove <- get_LOD_to_remove(params[["dat"]],
                                   params[["filtering_threshold"]]/100)

if(length(LOD_to_remove) == 0){
  cat("There were no metabolites with missing values ratio above the threshold.  \n   \n") 
}else{
  cat("Metabolites with missing values ratio above the threshold",
      ifelse(is.null(attr(params[["dat"]], "group")),
             ": ",
             " in each level of the group: "),
      paste0(LOD_to_remove, collapse = ", "),
      ".  \n   \n", sep = "")
}
```

```{r, results='asis'}
if(is.null(attr(params[["dat"]], "removed")[["LOD"]])){
  cat("No metabolite was removed based on the missing values ratio.  \n")
}else{
  cat("Metabolites removed based on the missing values ratio: ",
      paste0(attr(params[["dat"]], "removed")[["LOD"]], collapse = ", "),
      ".  \n", sep = "")
}
```

```{r, fig.height=4, fig.cap="Missing values ratios in each metabolite (up to 10 metabolites wiht the highest ratios are visible); (a) total ratios; (b) ratios of missing values of each type.", fig.pos="H", fig.align='center'}
NA_perc_plt <- plot_NA_percent(params[["dat"]],
                               interactive = FALSE)

y_levels <- ggplot_build(NA_perc_plt)$layout$panel_params[[1]]$y$breaks
cropped_lvls <- y_levels[(max(length(y_levels)-9, 1)):length(y_levels)]

NA_perc_plt_crop <- NA_perc_plt +
  ylim(cropped_lvls)

NA_perc_type_plt_crop <-
  plot_NA_percent(params[["dat"]],
                  type = "NA_type",
                  interactive = FALSE) +
  ylim(cropped_lvls)

ggarrange(NA_perc_plt_crop, NA_perc_type_plt_crop, ncol = 1, labels = "auto")
```

```{r, fig.height=4, fig.width=4, eval=((params[["filtering_threshold_ex"]]) & !is.null(attr(params[["dat"]], "group"))), fig.cap="Venn diagram of metabolites with missing values ratio greater than the threshold in any group level.", fig.pos="H", fig.align='center'}
if(lvls_count < 5)
  create_venn_diagram(params[["dat"]], params[["filtering_threshold"]]/100)
```

## Gaps completing

```{r, results='asis'}
if(!is.null(attr(params[["dat"]], "completed"))){
  cat("Data was completed using the following methods:  \n  \n")
  cat("* LOD method: ", params[["LOD_method"]], ",  \n", sep = "")
  cat("* LOD type: ", params[["LOD_type"]], ",  \n", sep = "")
  cat("* LLOQ method: ", params[["LLOQ_method"]], ",  \n", sep = "")
  cat("* ULOQ method: ", params[["ULOQ_method"]], ".  \n", sep = "")
}else{
  cat("Imputation was not performed on the data.")
}
```

```{r, fig.height=4, fig.width=6, fig.cap="Heatmap of missing values for each plate bar code (up to 10 first metabolites from the dataset are visible).", fig.pos="H", fig.align='center'}
aval_pb_codes <- params[["dat"]] %>%
  filter(`sample type` == "Sample") %>%
  select(`plate bar code`) %>%
  unlist() %>%
  unique()

aval_metabolites <- setdiff(attr(params[["dat"]], "metabolites"),
                            unlist(attr(params[["dat"]], "removed")))

heatmap_plts <- lapply(aval_pb_codes, function(pb_code){
  plot_heatmap(params[["dat"]], pb_code,
               include_title = TRUE) +
    ylim(aval_metabolites[1:min(10, length(aval_metabolites))])
})

NA_perc_plt_crop <- NA_perc_plt +
  ylim(cropped_lvls)

ggarrange(plotlist = heatmap_plts, ncol = min(length(heatmap_plts), 2))
```

```{r, results='asis', eval=!is.null(attr(params[["dat"]], "cv"))}
cat("## Quality control  \n  \n")
cat("CV threshold was set to ",
    params[["cv_threshold"]],
    "%.  \n   \n", sep = "")

CV_to_remove <- get_CV_to_remove(params[["dat"]],
                                 threshold = params[["cv_threshold"]]/100)

if(length(CV_to_remove) == 0){
  cat("There were no metabolites with CV greater than the threshold.  \n   \n")
}else{
  cat("Metabolites with CV greater than the threshold: ",
      paste0(CV_to_remove, collapse = ", "),
      ".  \n   \n", sep = "")
}

if(!is.null(attr(params[["dat"]], "removed")[["QC"]])){
  cat("Metabolites removed based on the CV: ",
      paste0(attr(params[["dat"]], "removed")[["QC"]],
             collapse = ", "),
      ".  \n", sep = "")
}else cat("No metabolites were removed based on the CV.  \n")
```

```{r, results='asis', eval=!is.null(attr(params[["dat"]], "cv"))}
cat("## Outlier detection   \n   \n
    For the biplot based on the PCA including all sample types, the threshold
    was set to ",
    params[["sample_type_PCA_threshold"]],
    "%. ", sep = "")
cat("For the plot of variance explained by each principal component, the
    threshold was set to ",
    params[["sample_type_PCA_variance_threshold"]],
    "%. ", sep = "")
cat("The maximum number of visible principal components was set to ",
    params[["sample_type_PCA_variance_max_num"]],
    ".", sep = "")
```

```{r, eval=!is.null(attr(params[["dat"]], "cv")), fig.height = 7, fig.width = 5, fig.cap="Principal Component Analysis (PCA) including all sample types; (a) scatter plot with colors indicating sample types; (b) biplot  showing similarities among compounds based on their contribution to variability; (c) barplot showing the proprotion of variance explained by each principal component.", fig.align='center', fig.pos="H"}
sample_pca_plt <- create_PCA_plot(
  params[["dat"]],
  group_by = "sample_type",
  type = "scatterplot",
  types_to_display = params[["sample_type_PCA_types"]],
  interactive = FALSE)

sample_biplot <- create_PCA_plot(params[["dat"]],
                          type = "biplot",
                          group_by = "sample_type",
                          threshold = params[["sample_type_PCA_threshold"]]/100,
                          types_to_display = params[["sample_type_PCA_types"]],
                          interactive = FALSE)

sample_pca_variance_plt <- pca_variance(params[["dat"]],
                                 group_by = "sample_type",
                                 threshold = params[["sample_type_PCA_variance_threshold"]]/100,
                                 max_num = params[["sample_type_PCA_variance_max_num"]],
                                 cumulative = params[["sample_type_PCA_variance_cum"]])

ggarrange(sample_pca_plt, sample_biplot, sample_pca_variance_plt,
          labels = "auto", ncol = 1)
```

```{r, results = "asis", eval=(!is.null(attr(params[["dat"]], "cv")) & !is.null(attr(params[["dat"]], "group")))}
cat("For the biplot based on the PCA including only samples, the threshold
    was set to ",
    params[["group_PCA_threshold"]],
    "%. ", sep = "")
cat("For the plot of variance explained by each principal component, the
    threshold was set to ",
    params[["group_PCA_variance_threshold"]],
    "%. ", sep = "")
cat("The maximum number of visible principal components was set to ",
    params[["group_PCA_variance_max_num"]],
    ".", sep = "")
```

```{r, eval=(!is.null(attr(params[["dat"]], "cv")) & !is.null(attr(params[["dat"]], "group"))), fig.height=7, fig.width=5, fig.cap="Principal Component Analysis (PCA) including only samples; (a) scatter plot with colors indicating group levels; (b) biplot  showing similarities among compounds based on their contribution to variability; (c) barplot shwoing the proprotion of variance explained by each principal component.", fig.align='center', fig.pos="H"}
group_pca_plt <- create_PCA_plot(
  params[["dat"]],
  group_by = "group",
  type = "scatterplot",
  interactive = FALSE)

group_biplot <- create_PCA_plot(params[["dat"]],
                          type = "biplot",
                          group_by = "group",
                          threshold = params[["group_PCA_threshold"]]/100,
                          interactive = FALSE)

group_pca_variance_plt <- pca_variance(params[["dat"]],
                                 group_by = "group",
                                 threshold = params[["group_PCA_variance_threshold"]]/100,
                                 max_num = params[["group_PCA_variance_max_num"]],
                                 cumulative = params[["group_PCA_variance_cum"]])

ggarrange(group_pca_plt, group_biplot, group_pca_variance_plt,
          labels = "auto", ncol = 1)
```





